# Exploratory Data Analysis - Air Pollution Example


### Goals

* Assess the quantity, quality of your data.
* Is there structure in the data that needs to be modeled?
* Which features are relevant for the mode?


```{r}
# import all libraries used in this notebook
require(sf)
require(geojsonio)
require(ggplot2)
require(dplyr)
suppressMessages(library(sf))
suppressMessages(library(geojsonio))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
options(width = 100)
```

## Dataset: Exposure to particulate matter less than 2.5 microns in diameter


Our goal is to build a model of ground-level air pollution, using measurements from ground monitors together
with satellite imaging data.
The dataset is taken from github repository: https://github.com/jgabry/bayes-vis-paper
which contain all materials for the paper **_Visualization in Bayesian workflow_**:    

Gabry, J. , Simpson, D. , Vehtari, A. , Betancourt, M. and Gelman, A. (2019),     
Visualization in Bayesian workflow. _J. R. Stat. Soc. A_, 182: 389-402. doi:10.1111/rssa.12378

* Published JRSS version: https://rss.onlinelibrary.wiley.com/doi/full/10.1111/rssa.12378
* arXiv preprint: https://arxiv.org/pdf/1709.01449.pdf (includes Supplementary Materials in appendix)

A local version of the raw data for this paper is in the `data` directory, named `GM_spdf.RData`.
This [RDATA file](https://github.com/jgabry/bayes-vis-paper/blob/master/bayes-vis.RData) contains a single R [SpatialPointsDataFrame](https://rdrr.io/cran/sp/man/SpatialPoints.html).

We have converted this to a [GeoJson](https://geojson.org) file, `data/air_spdf.geojson` which contains just the information needed for this analysis.

The dataset for this paper consists of measurements of ambient ${PM}_{2.5}$ collected from ground monitors
together with the corresponding measurement from high-resolution satellite data for that geo-location.
The data was collected accros 100 countries, which have been grouped into 7 categories by the World Health Organization (WHO).
Each observation consists of the following:

- 3-letter ISO country code
- location name
- a pair of latitude, longitude coordinates (in WGS84)
- ${PM}_{2.5}$ measurements from ground monitors, and log(ground PM)
- ${PM}_{2.5}$ measurements from via high-resolution satellite data, and log(satelite PM)
- World Health Organization (WHO) super-region name
- 6-component hierarchical clustering of ground monitor measurements
- 6-component hierarchical clustering of log ground monitor measurements


```{r}
pm_gdf = geojson_read("data/air_spdf.geojson", parse=TRUE, what="sp", stringsAsFactors=TRUE)
sites <- st_as_sf(pm_gdf)
sites$super_region_name <- factor(sites$super_region_name)
str(sites)
```


```{r}
head(sites)
```

What is the range of observed measurements of ${PM}_{2.5}$ and log ${PM}_{2.5}$ from ground monitors and satellites?


```{r}
cat("pm_25")
summary(sites$pm25)
cat("log_pm")
summary(sites$log_pm25)
cat("sat_2014")
summary(sites$sat_2014)
cat("log_sat")
summary(sites$log_sat)
```

To properly visualize the distribution of measurements around the globe we need a map which contains country boundaries.
We have downloaded a world map shapefile (in coordinate system WGS84) from
[Natural Earth](https://www.naturalearthdata.com/), which provides free public domain datasets.

We add the per-country WHO super region names and cluster regions to the world map so that we can color countries by super region and clustered region.


```{r}
world_map_gdf = st_read(dsn = "data/ne_110m_admin_0_countries", layer="ne_110m_admin_0_countries")
countries <- as.data.frame(sites) %>% select(iso3, super_region_name, super_region, cluster_region, cluster_log_region) %>% distinct()
world_map_plus_gdf <- merge(world_map_gdf, countries, by.x = 'ISO_A3_EH', by.y = 'iso3', all.x = TRUE)

options(repr.plot.width=20, repr.plot.height=10)
ggplot() +  
 geom_sf(data = world_map_plus_gdf, aes(fill=factor(super_region_name)), alpha=0.1) +
 geom_sf(data = sites, aes(color=factor(super_region_name))) +
 scale_color_manual(values = c('red', 'orange', 'magenta', 'green', 'purple', 'blue', 'cyan')) +
 scale_fill_manual(values = c('red', 'orange', 'magenta', 'green', 'purple', 'blue', 'cyan', 'ivory')) +
         guides(fill='none')
```

Alternatively, we can color countries by region and show ground monitor points by log_pm25 readings.   Green and yellows indicate low readings, red and orange indicate high readings.


```{r}
options(repr.plot.width=20, repr.plot.height=10)
ggplot() +  
 geom_sf(data = world_map_plus_gdf, aes(fill=factor(super_region_name)), alpha=0.2) +
 geom_sf(data = sites, aes(color=log_pm25)) +
 scale_color_gradient2(low='green', mid='yellow', high='red', midpoint=2.8) +
 scale_fill_manual(values = c('red', 'orange', 'magenta', 'green', 'purple', 'blue', 'cyan', 'ivory'))
```

## Data structure

The data has two levels of grouping:  low-level grouping by country (107 countries), and high-level grouping by super-region (7 categories) or cluster-region (6 clusters).

### Ground monitors per region


```{r}
options(repr.plot.width=10, repr.plot.height=5)
ggplot() +
    geom_bar(data=sites, aes(x=super_region, fill=factor(super_region_name))) +
    scale_fill_manual(values = c('red', 'orange', 'magenta', 'green', 'purple', 'blue', 'cyan')) +
    ggtitle('Ground Monitors per Region') + xlab('') + labs(fill='Super Region')
```

### Ground monitors per country

The number of per-country ground monitors depends on the size, overall development, and population density of the country.  China, the US, and India dominate, followed by other high income regions.


```{r}
sites %>%
  group_by(iso3) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  head(5)
```


```{r}
counts = sites %>%
  select(iso3) %>%
  count(iso3)

print(summary(counts$n))
```

Of these 107 countries surveyed, the median number of monitors per country is 9. We can visualize this distribution as a bar plot. Is this useful?


```{r}
ggplot() +
    geom_bar(data=sites, aes(x=iso3, fill=factor(super_region_name))) +
    scale_fill_manual(values = c('red', 'orange', 'magenta', 'green', 'purple', 'blue', 'cyan')) +
    ggtitle('Ground Monitors per Country') + xlab('region') + labs(fill='Super Region')
```

## Which features to use in the model?

Gabry et al consider 3 alternatives:

* Complete pooling - regress ground monitor readings on satellite data; all sites are the same.
* Partial pooling by WHO super region
* Partial pooling based on hierachical clustering of country-level average of ground monitor readings
   + here we also consider clustering on average of *log* ground monitor reading


### Cluster by ground monitor ${PM}_{2.5}$

What are the groupings for hierarhical clustering by ground monitor ${PM}_{2.5}$?


```{r}
cat("Sites per cluster_region")
table(sites$cluster_region)
```


```{r}
ggplot() +
    geom_bar(data=sites, aes(x=factor(cluster_region), fill=factor(cluster_region)))
```


```{r}
options(repr.plot.width=20, repr.plot.height=10)
ggplot() +  
 geom_sf(data = world_map_plus_gdf, aes(fill=factor(cluster_region)))
```


```{r}
options(repr.plot.width=20, repr.plot.height=10)
ggplot() +  
 geom_sf(data = world_map_plus_gdf, aes(fill=factor(cluster_region)), alpha=0.7) +
 geom_sf(data = sites, aes(color=log_pm25), alpha=0.7) +
 scale_color_gradient2(low='green', mid='yellow', high='red', midpoint=2.8)
```

### Cluster by log of ground monitor ${PM}_{2.5}$


```{r}
cat("Sites per cluster_log_region")
table(sites$cluster_log_region)
```


```{r}
options(repr.plot.width=10, repr.plot.height=5)
ggplot() +
    geom_bar(data=sites, aes(x=factor(cluster_log_region), fill=factor(cluster_log_region)))
```


```{r}
options(repr.plot.width=20, repr.plot.height=10)
ggplot() +  
 geom_sf(data = world_map_plus_gdf, aes(fill=factor(cluster_log_region)), alpha=0.7)
```


```{r}
ggplot() +  
 geom_sf(data = world_map_plus_gdf, aes(fill=factor(cluster_log_region)), alpha=0.7) +
 geom_sf(data = sites, aes(color=log_pm25), alpha=0.7) +
 scale_color_gradient2(low='green', mid='yellow', high='red', midpoint=2.8)
```
